---
title: ggplot basics
author: "Oleksii Yehorchenkov"
date: "`r Sys.Date()`"
output: html_document
---

```{r, results='hide', message=FALSE}
library(tidyverse)
library(ggrepel) # repel overlapping text labels
```

This tutorial is based on [R graphics with ggplot2 workshop notes](https://rpubs.com/phela51/471711)

## Points (scatterplot)

Now that we know about geometric objects and aesthetic mapping, 
we can make a `ggplot()`. `geom_point()` requires mappings for x 
and y, all others are optional.

**Example data: housing prices**

Let's look at housing prices.

```{r}
housing <- read_csv("./data/landdata-states.csv")
head(housing[1:5]) # view first 5 columns
```

create a subset for 1st quarter 2001

```{r}
hp2001Q1 <- housing |> filter(Date == 2001.25)
```

**Step 1:** create a blank canvas by specifying data:

```{r}
ggplot(data = hp2001Q1)
```

**Step 2:** specify aesthetic mappings 
(how you want to map variables to visual aspects):

here we map `Land_Value` and `Structure_Cost` to the x- and y-axes.

```{r}
ggplot(data = hp2001Q1, 
       mapping = aes(x = Land_Value, y = Structure_Cost))
```

**Step 3:** add new layers of geometric objects that will show up on the plot: 

here we use `geom_point()` to add a layer with point (dot) elements 
as the geometric shapes to represent the data.

```{r}
ggplot(data = hp2001Q1, aes(x = Land_Value, y = Structure_Cost)) +
    geom_point()
```

## Lines (prediction line)

A plot constructed with `ggplot()` can have more than one geom. 
In that case the mappings established in the `ggplot()` 
call are plot defaults that can be added to or overridden --- 
this is referred to as **aesthetic inheritance**. 
Our plot could use a regression line:

get predicted values from a linear regression

```{r}
hp2001Q1$pred_SC <- lm(Structure_Cost ~ log(Land_Value), 
                       data = hp2001Q1) |> predict()

p1 <- ggplot(hp2001Q1, aes(x = log(Land_Value), y = Structure_Cost))

p1 + geom_point(aes(color = Home_Value)) + # values for x and y are inherited from the ggplot() call above
    geom_line(aes(y = pred_SC)) # add predicted values to the plot overriding the y values from the ggplot() call above
```

## Smoothers

Not all geometric objects are simple shapes; 
the smooth geom includes a line and a ribbon.

```{r}
p1 +
    geom_point(aes(color = Home_Value)) +
    geom_smooth()
```

## Text (label points)

Each geom accepts a particular set of mappings; 
for example `geom_text()` accepts a `label` mapping.

```{r}
p1 + 
    geom_text(aes(label = State), size = 3)
```

But what if we want to include points and labels? 
We can use `geom_text_repel()` to keep labels from overlapping 
the points and each other.

```{r}
p1 + 
    geom_point() + 
    geom_text_repel(aes(label = State), size = 3)
```

## Aesthetic mapping VS assignment

1.  Variables are **mapped** to aesthetics inside the `aes()` function.

```{r}
p1 +
    geom_point(aes(size = Home_Value))
```

2.  Constants are **assigned** to aesthetics outside the `aes()` call

```{r}
p1 +
    geom_point(size = 2)
```

This sometimes leads to confusion, as in this example:

```{r}
p1 +
    geom_point(aes(size = 2),# incorrect! 2 is not a variable
               color="red") # this is fine -- all points red
```

## Mapping variables to other aesthetics

Other aesthetics are mapped in the same way as x and y 
in the previous example.

```{r}
p1 +
    geom_point(aes(color = Home_Value, shape = region))
```

## Other visualizations

### Histogramm

A histogram is an approximate representation of the distribution of numerical data. To construct a histogram, the first step is to "bin" (or "bucket") the range of values— divide the entire range of values into a series of intervals—and then count how many values fall into each interval. The bins are usually specified as consecutive, non-overlapping intervals of a variable. The bins (intervals) must be adjacent and are often (but not required to be) of equal size.

```{r}
p2 <- ggplot(housing, aes(x = Home_Value))
p2 + geom_histogram(bins=30)
```

We can use a density estimate, which is a smoothed version of the histogram. This is a useful alternative to the histogram for continuous data that comes from an underlying smooth distribution

```{r}
p2 <- ggplot(housing, aes(x = Home_Value))
p2 + geom_density(color="blue", linewidth=2)
```

### Boxplot

Boxlots are a type of data visualization that shows summary statistics for data.

More specifically, boxplots visualize what we call the “five number summary.” The five number summary is a set of values that includes:

- the minimum (-1.5 * IRQ)
- the first quartile (25th percentile)
- the median
- the third quartile (75th percentile)
- the maximum (+1.5 * IRQ)
- outliers

```{r}
p3 <- ggplot(housing, aes(region, Home_Value))
p3 + geom_boxplot()
```

